%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                      %
%     File: Thesis_Acknowledgments.tex                                 %
%     Tex Master: Thesis.tex                                           %
%                                                                      %
%     Author: Andre C. Marta                                           %
%     Last modified : 21 Jan 2011                                      %
%                                                                      %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section*{\acknowledgments}

% Add entry in the table of contents as section
\addcontentsline{toc}{section}{\acknowledgments}

%Supervisors
It would come as a great oversight if I was not aware of the help and contribution I received for the completion of this undertaking.

I should start by expressing deep appreciation to my supervisors, Dr. Ana Fred and Dr. Helena Lopes.
The freedom I was allowed was crucial for keeping my motivation up for such a long period.
When that was lacking, their guidance and encouragement quickly put me back on track.

%Family & Friends
I leave my warmest regards to the family I built throughout my military and academic career, during these last years.
Your camaraderie has shaped where I stand and who I am today.
Your support throughout these long months was invaluable.

% Bella
To my loving girlfriend, who would always hear my boundless enthusiasm or disheartening frustrations, whichever was the case, and who followed me closest in the last months, I am deeply grateful.
You motivated me most of all, and that was no small task.

And to my parents, thank you for all your support and motivation during all these years.
You were always caring and encouraging and I am forever grateful for that.

Last, but not least, thank you to all my friends and family for understanding my absence during these challenging months.
\cleardoublepage

